---
title: "Financial Report"
author: "Ghanshyam Thakur"
date: "2026-02-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Scenario: 
You are a Data Scientist working for a consulting firm. One of your colleagues 
from the Auditing department has asked you to help them assess the financial statement of 
organization X. 
You have been supplied with two vectors of data: monthly revenue and monthly expenses for 
the financial year in question. Your task is to calculate the following financial metrics: - 
profit for each month - - - - - - 
profit after tax for each month (the tax rate is 30%) 
profit margin for each month - equals to profit after tax divided by revenue 
good months - where the profit after tax was greater than the mean for the year 
bad months - where the profit after tax was less than the mean for the year 
the best month - where the profit after tax was max for the year 
the worst month - where the profit after tax was min for the year 
All results need to be presented as vectors. 
Results for dollar values need to be calculated with $0.01 precision, but need to be presented 
in Units of $1,000 (i.e. 1k) with no decimal points. 
Results for the profit margin ratio need to be presented in units of % with no decimal points. 
Note: You colleague has warned you that it is okay for tax for any given month to be 
negative (in accounting terms, negative tax translates into a deferred tax asset). 


## Data and Methodology
In this section, we initialize the monthly revenue and expense data. We calculate the profit, account for a 30% tax rate, and determine the profit margins.

```{r calculations}
# Data Input
revenue <- c(14574.49, 7606.46, 8611.41, 9175.41, 8058.65, 8105.44, 11496.28, 9766.09, 10305.32, 14379.96, 10713.97, 15433.50)
expenses <- c(12051.82, 5695.07, 12319.20, 12089.72, 8658.57, 840.20, 3285.73, 5821.12, 6976.93, 16618.61, 10054.37, 3803.96)
```


```{}
# Profit and Tax Calculations
profit <- revenue - expenses
tax <- round(profit * 0.3, 2)
profit.after.tax <- profit - tax
```


### Output
```{}
profit.after.tax  
 1765.87  1337.97 -2595.45 -2040.02  -419.94  5085.67  5747.38 2761.48  2329.87 -1567.05   461.72  8140.68
 ```
 **Net Profitability**: You ended the year with a total net profit of $21,225.27 after taxes.

**The "V" Recovery:** After a rough start and a heavy loss in Month 3 (your worst month), the business saw a massive turnaround, peaking in Month 12 (your best month).

**Volatility:** The business is inconsistent. You had 4 months of losses, but the profitable months (especially 6, 7, and 12) were strong enough to more than cover the deficits.

**Tax Impact:** The 30% tax rate took about $9,096 out of your raw profits, but the "tax credits" from loss months helped soften the blow to your bottom line.

```{}
# profit margin for each month - equals to profit after tax divided by revenue

profit.margin = profit.after.tax / revenue

profit.margin.per = round(profit.margin,2)*100
profit.margin.per
```
### Output
```{}
profit.margin.per
 [1]  12  18 -30 -22  -5  63  50  28  23 -11   4  53
 ```
 
 
**Best Efficiency:** Month 6 is your most efficient month ($63\%$), followed by Month 12 ($53\%$) and Month 7 ($50\%$).

**Worst Efficiency:** Month 3 ($ -30%$) and Month 4 ($-22%$) show significant capital drain relative to revenue.

**Overall Trend:** Your margins are highly unstable, swinging from a loss of $30\%$ to a gain of $63\%$. This suggests that while your sales are steady, your cost control or operational efficiency varies wildly throughout the year.


```{}
#good months - where the profit after tax was greater than the mean for the year

mean.pet = mean(profit.after.tax)
mean.pet
```
### Output
```{}
good.months  
 [1]  TRUE FALSE FALSE FALSE FALSE  TRUE  TRUE  TRUE  TRUE FALSE FALSE
[12]  TRUE
```
**Above-Average Performance:** Six out of the twelve months (Months 1, 6, 7, 8, 9, and 12) performed better than the yearly average.

**Strong Consistency in Mid-to-Late Year:** A solid "good" streak occurred from Month 6 through Month 9, indicating a period of sustained high performance during the middle of the year.

**Month 12 Peak:** Month 12 remains the standout, significantly exceeding the average and lifting the overall annual mean.

```{}
#bad months - where the profit after tax was less than the mean for the year

bad.months = profit.after.tax < mean.pet
bad.months
```
### Output
```{}
bad.months
FALSE  TRUE  TRUE  TRUE  TRUE FALSE FALSE FALSE FALSE  TRUE  TRUE FALSE
```

**Frequency:** You had 6 bad months that fell below the annual average of 1750.68.

**The "Danger Zone":** A continuous four-month stretch (Months 2, 3, 4, and 5) underperformed, which likely strained cash flow during that period.

**Late-Year Dip:** Despite a strong recovery in the middle of the year, Months 10 and 11 also fell into the "bad" category before the final peak in Month 12.

```{}
#the best month - where the profit after tax was max for the year 
best.month = max(profit.after.tax)
best.month
profit.after.tax
```
### Output

```{}
best.month
8140.68
```
**Peak Performance:** Month 12 was the most successful month of the year, yielding the highest post-tax profit.
**Impact:** This single month's performance was so strong that it significantly boosted the annual average and helped offset the losses incurred earlier in the year.
**Efficiency:** Combined with the earlier calculation, this month also boasted a high profit margin of $53\%$. 

```{}
#the worst month - where the profit after tax was min for the year.
worse.month = min(profit.after.tax)
worse.month
```
output
```{}
worse.month
[1] -2595.45
```
**Lowest Point:** Month 3 was your most challenging period, resulting in the year's largest net loss.
**Significance:** This represents a significant outlier where expenses outpaced revenue by a wide margin, leading to a negative $30\%$ profit margin.
**Cash Flow Strain:** This month accounts for the steepest decline in the "bad months" streak (Months 2–5).




This code creates a consolidated matrix g that provides an overview of the entire year's performance. Because the matrix contains a mix of data types, R converts the logical values (TRUE/FALSE) into numbers ($1$ and $0$).

```{}
g <- rbind(
  revenue.1000,
  expenses.1000,
  profit.1000,
  profit.after.tax.1000,
  profit.margin,
  good.months,
  bad.months,
  best.month,
  worse.month
)
g
```

## output



```{}
revenue.1000             15.0000000     8.0000000     9.0000000   
expenses.1000            12.0000000     6.0000000    12.0000000
profit.1000               3.0000000     2.0000000    -4.0000000
profit.after.tex.1000     2.0000000     1.0000000    -3.0000000
profit.margin             0.1211617     0.1758992    -0.3013966
good.months               1.0000000     0.0000000     0.0000000
bad.months                0.0000000     1.0000000     1.0000000
best.month             8140.6800000  8140.6800000  8140.6800000
worse.month           -2595.4500000 -2595.4500000 -2595.4500000
                               [,4]          [,5]          [,6]
revenue.1000              9.0000000  8.000000e+00     8.0000000
expenses.1000            12.0000000  9.000000e+00     1.0000000
profit.1000              -3.0000000 -1.000000e+00     7.0000000
profit.after.tex.1000    -2.0000000  0.000000e+00     5.0000000
profit.margin            -0.2223356 -5.211047e-02     0.6274391
good.months               0.0000000  0.000000e+00     1.0000000
bad.months                1.0000000  1.000000e+00     0.0000000
best.month             8140.6800000  8.140680e+03  8140.6800000
worse.month           -2595.4500000 -2.595450e+03 -2595.4500000
                               [,7]          [,8]          [,9]
revenue.1000             11.0000000    10.0000000    10.0000000
expenses.1000             3.0000000     6.0000000     7.0000000
profit.1000               8.0000000     4.0000000     3.0000000
profit.after.tex.1000     6.0000000     3.0000000     2.0000000
profit.margin             0.4999339     0.2827621     0.2260842
good.months               1.0000000     1.0000000     1.0000000
bad.months                0.0000000     0.0000000     0.0000000
best.month             8140.6800000  8140.6800000  8140.6800000
worse.month           -2595.4500000 -2595.4500000 -2595.4500000
                              [,10]         [,11]         [,12]
revenue.1000             14.0000000  1.100000e+01    15.0000000
expenses.1000            17.0000000  1.000000e+01     4.0000000
profit.1000              -2.0000000  1.000000e+00    12.0000000
profit.after.tex.1000    -2.0000000  0.000000e+00     8.0000000
profit.margin            -0.1089746  4.309514e-02     0.5274682
good.months               0.0000000  0.000000e+00     1.0000000
bad.months                1.0000000  1.000000e+00     0.0000000
best.month             8140.6800000  8.140680e+03  8140.6800000
worse.month           -2595.4500000 -2.595450e+03 -2595.4500000
```


## Interpretation of Matrix g
The matrix allows you to compare different financial metrics side-by-side across all 12 months:

**Overview of Cash Flow (Units in $1,000s$):**The first four rows (revenue.1000 through profit.after.tax.1000) give you a simplified look at the scale of the business. For example, in Month 12, you generated $\$15,000$ in revenue and kept $\$8,000$ as net profit.

**Efficiency Trends (profit.margin):**This row tracks how much of every dollar earned was kept. It highlights the mid-year efficiency peak where margins hit $63\%$ (Month 6) and $50\%$ (Month 7).

**Performance Flags (Binary 1/0):** good.months / bad.months: These rows quickly categorize each month against the annual average. You can see a clear "green" streak from Month 6 to Month 9.best.month: Marked with a $1$ only in Month 12, confirming it as the yearly high point.worse.month: Marked with a $1$ only in Month 3, confirming it as the yearly low point.

### Summary 
Insight By binding these vectors, we  created a dashboard. It shows that while the business has high-revenue months (like Month 10 with $\$14k$), revenue alone doesn't guarantee success; Month 10 was actually a bad month ($0$ in good.months) because expenses were too high. The most successful months were characterized by either high revenue with controlled costs (Month 12) or average revenue with exceptionally low costs (Month 6).



### 1. The Bottom Line

Despite several periods of loss, the year was a success. business generated a total **Net Profit (After Tax) of $21,225.27**, maintaining an average monthly profit of about **$1,751**.

### 2. Highs and Lows

* **Best Month (Month 12):** The strongest performance, contributing **$8,140** to the bottom line with a **53% margin**.
* **Worst Month (Month 3):** The deepest deficit, losing **$2,595** with a negative **30% margin**.
* **The Efficiency Peak (Month 6):** Even with average revenue, this was your most "lean" month with a massive **63% profit margin**.

### 3. Key Takeaway

business has a **50/50 split** between "Good" and "Bad" months. Revenue is relatively stable, but **expenses are unpredictable**. The high-profit months (6, 7, and 12) act as the "engine" that carries the business through the heavy spending spikes seen in the first half of the year and Month 10.